COSC3500 \\ 2D Orbital Simulation Report

### 2018-09-21 reports/ms1_report.tex
Init
Big refactor

>>>>

\section*{Parallelism}
A description of your parallel implementation. How do you make use of OpenMP and MPI? How is information and work shared between threads and processes.  Any optimisations applied and/or challenges encountered will go here.
\subsection*{Implementation}
\subsubsection*{OpenMP}

I slapped as many \texttt{\#pragma omp parallel for shared(bodies)} on as many loops as I could find, and quickly discovered massive lock contention issues associated with \texttt{\#pragma omp critical}. Pairwise force updates between bodies:

\begin{verbatim}
    for (size_t i = 0; i < bodies.size() - 1; i++) {
        auto& x = bodies[i];

        for (size_t j = i + 1; j < bodies.size(); j++) {
            auto& y = bodies[j];
                x.exert_force_bidirectionally(y);
        }
    }
\end{verbatim}

where

\begin{verbatim}
void Body::exert_force_bidirectionally(Body& there) {
    ...
    #pragma omp critical
    {
    Fx += delta_x * scale_factor;
    Fy += delta_y * scale_factor;

    there.Fx -= delta_x * scale_factor;
    there.Fy -= delta_y * scale_factor;
    }
}
\end{verbatim}

which had provided massive performance gains in the serial implementation over unidirectional updates was now causing massive parallel slowdowns, where increasing the \texttt{OMP\_NUM\_THREADS} caused a proportional increase in walltime. Instead, a \texttt{critical}-less comparison schema was devised,

\begin{verbatim}
#pragma omp parallel for shared(bodies)
for (size_t i = 0; i < bodies.size(); i++) {
    auto& x = bodies[i];

    for (size_t j = 0; j < bodies.size(); j++) {
        auto& y = bodies[j];

        if (&bodies[i] != &bodies[j]) {
            // XXX: Do NOT swap these around. You will cause
            // race conditions
            x.exert_force_unidirectionally(y);
        }
    }
}
\end{verbatim}

that provided the sought after parallel speedup. This parallel speedup was found to be present at where $n$-bodies was high. Testing focussed primarily on $n > 5000$,\footnote{In MS1, I failed to test simulations with a large number of $n$, as I was focussed on testing the accuracy of the implementation, and generated input files would cause high numbers of collisions that made accuracy assessment difficult. Testing $n=10000$, we find a 10x speedup with by using a serial Barnes-Hut implementation (Avg. $1.12$s, as opposed to $9.87$s without Barnes-Hut). Barnes-Hut afforded a much simpler parallelization scheme, as each body could calculate its force against the quadtree concurrently, without locks} where force and integration calculation overhead dominated thread creation overhead. Parallel speedup vanished around the $n \approx 500$ mark.

% Increasing \texttt{OMP_NUM_THREADS} beyond \texttt{4} failed to provide a speedup at $n = 10000$.

\subsubsection*{MPI}

Looking at our high-level implementation, the following two steps consume the vast majority of our computing power:

\begin{enumerate}
    \item Inserted all \texttt{Body}s into a fresh \texttt{QuadTree} on full timesteps, traversing the \texttt{QuadTree} with every \texttt{Body} to calculate forces ($O(n \log n)$)
    \item Performed the appropriate \textit{Leapfrog} step to derive new velocities \textit{or} positions
\end{enumerate}

The \textit{Leapfrog} step is the easiest step to distribute amongst many nodes, as it does not require the comparison of a body to all other bodies - each node can compute the step on a subset of all bodies, and later combine each subset with relatively simple MPI code.

This can be achieved with \texttt{MPI\_Scatterv} and \texttt{MPI\_Gatherv}
\footnote{These \texttt{v} variants support custom chunking sizes e.g. if we have 15 bodies we can send $4, 4, 4, 3$ bodies to each node respectively, which \texttt{MPI\_Scatter} and \texttt{MPI\_Gather} do not support dynamic chunking}. We decompose each dimension of our \texttt{Body}s into arrays, (e.g. the $x$ position, the $x$ veolocity, etc.), \texttt{MPI\_Scatterv} these arrays, recompose them into \texttt{Body}s, run \texttt{Leapfrog} on this subset on each node, decompose the subset into arrays, \texttt{MPI\_Gatherv} these arrays back to the root node, and recompose them into $\texttt{Bodys}$.

\begin{verbatim}
std::vector<Body> sbodies = scatter_bodies(bodies, size, rank);

#pragma omp parallel for shared(bodies)
for (size_t i = 0; i < sbodies.size(); i++) {
    auto& body = sbodies[i];

    if (step % 2 == LEAP) {
        body.leap(timestep);
    }
    else {
        body.frog(timestep);
    }
}

bodies = gather_bodies(sbodies, size, rank, bodies.size());
\end{verbatim}

where \texttt{sbodies} is our subset of bodies distributed to each node.

Because force calculation is only performed on the root node, this messaging architecture is \textit{master-slave}, where the root node coordinates gathers and scatters. Slave nodes, are, unfortunately, underutilized. Further work could be sent distributing force calculation work, doing away with a root node entirely, except for logging.

\includegraphics[width=3cm]{mpi_gather}
\includegraphics[width=3cm]{mpi_scatter}

% A description of how you have verified the correctness of your parallel implementation. (See comments for MS1)
\subsection*{Verification \& Correctness}

Because of my confidence in the correctness of the serial implementation, it served as a reference implementation for the verification of the parallel implementation.

Both implementations were run with identical input parameters, and their outputs were \texttt{diff}d. When run without MPI enabled, outputs were identical. When run with MPI, floating point abnormalities (initially affecting the least significant digits of each floating point value) gradually caused each diff to diverge. I hypothesize that the MPI serialization and deserialization procedure was truncating precision and causing this divergence. Perhaps some option exists to ensure that MPI guarantees floating point precision.

% A credible plan of jobs to run to measure the scalability of your parallel implementation. The idea here is that you could take this plan an start performing timing experiments. You should not organise your testing so that if your test plan is curtailed, then you can still draw sensible conclusions from the results you do get.
\subsection*{Scalability}

We have mulitple dimensions that we can change to measure the scalability of the parallel implementation.

\begin{enumerate}
    \item The number of bodies ($n$) (\texttt{512, 1024, 2048, 4096, 8192, 16384})
    \item \texttt{ENABLE\_BARNES\_HUT} (\texttt{true, false})
    \item \texttt{--nodes} (\texttt{1, 4, 16})
    \item \texttt{--ntasks} (\texttt{1, 4, 16})
    \item \texttt{--ntasks-per-node} (\texttt{1, 4})
    \item \texttt{--cpus-per-task} (\texttt{1, 4, 16})
\end{enumerate}

Running tests with all valid combinations of these variables, with a \texttt{numTimeSteps} as low as 8, would produce useful data, across multiple dimension, with minimal load on cluster. It would be unlikely that this plan would be curtailed.
